graphsim: An R package for simulating gene expression data from graph structures of biological pathways

### 2020-06-30 paper/paper.Rmd
increment version for major release
update manscript with feedback

---
title: 'graphsim: An R package for simulating gene expression data from graph structures of biological pathways'
output:
  rmarkdown::pdf_document:
    fig_crop: no
    keep_md: TRUE
    #keep_tex: TRUE
    fig_caption: yes
tags:
  - R
  - gene-expression
  - simulation
  - genomics
  - pathway
  - network
authors:
  - name: S. Thomas Kelly
    email: "tom.kelly@postgrad.otago.ac.nz, tom.kelly@riken.jp"
    orcid: 0000-0003-3904-6690
    affiliation: "1, 2" # (Multiple affiliations must be quoted)
  - name: Michael A. Black
    email: mik.black@otago.ac.nz
    orcid: 0000-0003-1174-6054
    affiliation: "1"
affiliations:
 - name: "Department of Biochemistry, University of Otago, PO Box 56, Dunedin 9054, New Zealand"
   index: 1
 - name: "RIKEN Center for Integrative Medical Sciences, Suehiro-cho-1-7-22, Tsurumi Ward, Yokohama, Kanagawa 230-0045, Japan"
   index: 2
date: "`r  format(Sys.time(), '%d %B %Y')`"
bibliography: paper.bib
header-includes:
  - \usepackage{caption}
---

@@

### Summary
Transcriptomic analysis is used to capture the molecular state of a cell
or sample in many biological and medical applications. In addition to 
identifying alterations in activity at the level of individual genes, 
understanding changes in the gene networks that regulate fundamental
biological mechanisms is also an important objective of molecular 
analysis. As a result, databases that describe biological pathways 
are increasingly relied on to assist with the interpretation of results
from large-scale genomics studies. Incorporating information from 
biological pathways and gene regulatory networks into a genomic data
analysis is a popular strategy, and there are many methods that provide
this functionality for gene expression data. When developing or comparing
such methods, it is important to gain an accurate assessment of their 
performance, with simulation-based validation studies a popular choice.
This necessitates the use of simulated data that correctly accounts for
pathway relationships and correlations. Here we present a versatile
statistical framework to simulate correlated gene expression data from
biological pathways, by sampling from a multivariate normal distribution
derived from a graph structure. This procedure has been released as the
\texttt{graphsim} R package on CRAN and GitHub (\url{https://github.com/TomKellyGenetics/graphsim})
 and is compatible with any graph structure that can be described using
the `igraph` package. This package allows the simulation of biological
pathways from a graph structure based on a statistical model of gene expression.

@@

Network analysis of molecular biological pathways has the potential to
lead to new insights into biology and medical genetics
[@Barabasi2004; @Hu2016]. Since gene expression profiles capture a
consistent signature of the regulatory state of a cell
[@Perou2000; @Ozsolak2011; @Svensson2018], they can be used to analyse
complex molecular states with genome-scale data. However, biological
pathways are often analysed in a reductionist paradigm as amorphous sets
of genes involved in particular functions, despite the fact that the
relationships defined by pathway structure could further inform gene
expression analyses. In many cases, the pathway relationships are
well-defined, experimentally-validated, and are available in public
databases [@Reactome]. As a result, network analysis techniques could
play an important role in furthering our understanding of biological
pathways and aiding in the interpretation of genomics studies.

Gene networks provide insights into how cells are regulated, by mapping
regulatory interactions between target genes and transcription factors,
enhancers, and sites of epigenetic marks or chromatin structures
[@Barabasi2004; @Yamaguchi2007]. Inference of these regulatory
interactions for genomics investigations has the potential to radically
expand the range of candidate biological pathways to be further
explored, or to improve the accuracy of bioinformatics and functional
genomic analysis. A number of methods have already been developed to
utilise timecourse gene expression data [@Arner2015; @Yamaguchi2007]
using gene regulatory modules in state-space models and recursive vector
autoregressive models [@Hirose2008; @Shimamura2009]. Various approaches
to gene regulation and networks at the genome-wide scale have led to
novel biological insights [@Arner2015; @Komatsu2013]. However, inference
of regulatory networks has thus far primarily relied on experimental
validation or resampling-based approaches to estimate the likelihood
of specific network modules being predicted [@Markowetz2007; @Hawe2019].

Simulated datasets with pathway structure are of particular interest for
benchmarking regulatory network inference techniques and methods
being developed for genomics data with complex biological 
interactions [@Schaffter2011; @Saelens2019]. Dynamical models using
differential equations have been employed, such as by GeneNetWeaver
[@Schaffter2011], to generate simulated datasets
specifically for benchmarking gene regulatory network inference techniques.
There is renewed interest in modelling biological pathways and simulating
data for benchmarking due to the emergence of single-cell genomics
technologies and the plethora of bioinformatics techniques developed
to use this data [@Zappia2017; @Saelens2019]. Packages such as 'splatter'
[@Zappia2017], which uses the gamma-poisson distribution,
have been developed to model single-cell data.
SERGIO [@Dibaeinia2019] and dyngen [@Cannoodt2020] build on
thisby adding gene regulatory networks and multimodality
respectively. These methods have been designed based on known
or synthetic reaction states and added stochasticity to deterministic
relationships. However, it is computationally-intensive to model
these reactions at scale over iterations in benchmarking.
In some cases, it is only necessary to model the statistical
variability and ``noise'' of RNA-Seq data to evaluate methods
with multivariate correlation structures.

There is a need, therefore, for a systematic framework for statistical
modelling and simulation of gene expression data derived from
hypothetical, inferred or known gene networks. Here we present a
package to achieve this, where samples from a multivariate normal
distribution are used to generate normally-distributed log-expression
data, with correlations between genes derived from the structure of the
underlying pathway or gene regulatory network. This methodology enables
simulation of expression profiles that approximate the log-transformed
and normalised data from microarray and bulk or single-cell RNA-Seq
experiments. This procedure has been released as the package to enable
the generation of simulated gene expression datasets containing pathway
relationships from a known underlying network. These simulated datasets
can be used to evaluate various bioinformatics methodologies, including
statistical and network inference procedures.

@@

Here we present a procedure to simulate gene expression data with
correlation structure derived from a known graph structure. This
procedure assumes that transcriptomic data have been generated and
follow a log-normal distribution (i.e.,
$log(X_{ij}) \sim MVN({\bf\mu}, \Sigma)$, where ${\bf\mu}$ and $\Sigma$
are the mean vector and variance-covariance matrix respectively, for
gene expression data derived from a biological pathway) after
appropriate normalisation [@Law2014; @Li2015]. Log-normality of gene
expression matches the assumptions of the popular \texttt{limma} package [@limma], which is
often used for the analysis of intensity-based data from gene expression
microarray studies and count-based data from RNA-Seq experiments. This
approach has also been applied for modelling UMI-based count data from
single-cell RNA-Seq experiments in the package [@Wang2018].

@@

Computing the nearest positive definite matrix is necessary to ensure
that the variance-covariance matrix could be inverted when used as a
parameter in multivariate normal simulations, particularly when negative
correlations are included for inhibitions (as shown below). Matrices
that could not be inverted occurred rarely with biologically plausible
graph structures but this approach allows for the computation of a
plausible correlation matrix when the given graph structure is
incomplete or contains loops. When required, the nearest positive
definite matrix is computed using the `nearPD` function of the \texttt{Matrix} R package
[@Matrix] to perform Higham's algorithm [@Higham2002] on
variance-covariance matrices. The \texttt{graphsim} package gives a warning when this
occurs.

@@

The graph structure in
Figure [1a](#fig:simple_graph:first){reference-type="ref"
reference="fig:simple_graph:first"} was used to simulate correlated gene
expression data by sampling from a multivariate normal distribution
using the R package [@Genz2009; @mvtnorm]. The graph structure
visualisation in
Figure [1](#fig:simple_graph){reference-type="ref"
reference="fig:simple_graph"} was specifically developed for (directed)
\texttt{igraph} objects in and is available in the and packages. The
\texttt{plot\_directed} function enables customisation of plot parameters for
each node or edge, and mixed (directed) edge types for indicating
activation or inhibition. These inhibition links (which occur frequently
in biological pathways) are demonstrated in
Figure [1b](#fig:simple_graph:second){reference-type="ref"
reference="fig:simple_graph:second"}.

@@

```{r, eval = FALSE}
#install packages required (once per machine)
install.packages("graphsim")
```
```{r, warning=FALSE, results='hide', message=FALSE}
#load required packages (once per R instance)
library("graphsim")
#load packages for examples
library("igraph"); library("gplots"); library("scales")
```

@@

```{r simple_graph, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.375\\linewidth', out.width='.375\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulated graph structures}. A constructed graph structure used as an example to demonstrate the simulation procedure in Figures 2 and 3. Activating links are denoted by black arrows and inhibiting links by red edges. Inhibiting edges have been highlighted in red.', echo=FALSE}
#generate graph structure
graph_edges <- rbind(c("A", "C"), c("B", "C"), c("C", "D"), c("D", "E"),
   c("D", "F"), c("F", "G"), c("F", "I"), c("H", "I"))
graph <- graph.edgelist(graph_edges, directed = TRUE)

@@

#simulated data
expr <- generate_expression(100, graph, cor = 0.8, mean = 0,
  comm = FALSE, dist =TRUE, absolute = FALSE, state = state)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none",
          col = colorpanel(50, "white", "red"),
  colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
  colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
```

```{r simulation_activating, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.375\\linewidth', out.width='.375\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from a graph structure}. An example of a graph structure (a) that has been used to derive a relationship matrix (b), $\\Sigma$  matrix (c) and correlation structure (d) from the relative distances between the nodes. Non-negative values are coloured white to red from $0$ to $1$. This $\\Sigma$ matrix has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here genes with closer relationships in the pathway structure show a higher correlation between simulated values.', warning=FALSE, echo=FALSE}
# activating graph
state <- rep(1, length(E(graph)))
plot_directed(graph, state=state, layout = layout.kamada.kawai,
  cex.node=2, cex.arrow=4, arrow_clip = 0.2)
mtext(text = "(a) Activating pathway structure", side=1, line=3.5, at=0.075, adj=0.5, cex=1.75)
box()
#plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
  scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(b) Relationship matrix", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, cor = 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#simulated data
expr <- generate_expression(100, graph, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE, state = state)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(d) Simulated correlation", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "samples", side=1, line=1.5, at=0.2, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=-0.4, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
```

```{r, include=FALSE, warning=FALSE}
pdf("Plotsimulation_activating-5.pdf", width = 12.14, height = 6.072)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "", mar = c(6, 6), keysize = 1)
mtext(text = "samples", side=1, line=1.5, at=0.55, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=0.45, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0.5, adj=0.5, cex=1.75)
dev.off()
#system("sed -i.bak '/Plotsimulation_activating-5/s/\\(.*\\)width=[.]415/\1width=.830/g' paper.md && rm paper.md.bak")
#system("sed -i.bak '/Plotsimulation_activating-5/s/\\(.*\\)width=[.]415/\1width=.830/g' paper.tex && rm paper.tex.bak")
#sed -i '/Plot.*-5/s/\(.*\)width=[.]415/\1width=.830/g' paper.md paper.tex
```

@@

```{r simulation_inhibiting, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.375\\linewidth', out.width='.375\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from graph structure with inhibitions}. An example of a graph structure (a), that has been used to derive a relationship matrix (b), $\\Sigma$ matrix (c), and correlation structure (d), from the relative distances between the nodes. These values are coloured blue to red from $-1$ to $1$. This has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here the inhibitory relationships between genes are reflected in negatively correlated simulated  values.', warning=FALSE, echo=FALSE}
#generate parameters for inhibitions
state <- c(1, 1, -1, 1, 1, 1, 1, -1)
plot_directed(graph, state=state, layout = layout.kamada.kawai,
  cex.node=2, cex.arrow=4, arrow_clip = 0.2)
mtext(text = "(a) Inhibiting pathway structure", side=1, line=3.5, at=0.075, adj=0.5, cex=1.75)
box()
#plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
  scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(b) Relationship matrix", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, state, cor = 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#simulated data
expr <- generate_expression(100, graph, state, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(d) Simulated correlation", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "samples", side=1, line=1.5, at=0.2, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=-0.4, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
```

```{r, include=FALSE, warning=FALSE}
pdf("Plotsimulation_inhibiting-5.pdf", width = 12.14, height = 6.072)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "", mar = c(6, 6), keysize = 1)
mtext(text = "samples", side=1, line=1.5, at=0.55, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=0.45, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0.5, adj=0.5, cex=1.75)
dev.off()
#("sed -i.bak '/Plotsimulation_inhibiting-5/s/\\(.*\\)width=[.]415/\1width=.830/g' paper.md && rm paper.md.bak")
#system("sed -i.bak '/Plotsimulation_inhibiting-5/s/\\(.*\\)width=[.]415/\1width=.830/g' paper.tex && rm paper.tex.bak")
#sed -i '/Plot.*-5/s/\(.*\)width=[.]415/\1width=.830/g' paper.md paper.tex
```

>>>>

---
title: 'graphsim: An R package for simulating gene expression data from graph structures of biological pathways'
output:
  rmarkdown::pdf_document:
    fig_crop: no
    #keep_md: TRUE
    keep_tex: TRUE
    fig_caption: yes
tags:
  - R
  - gene-expression
  - simulation
  - genomics
  - pathway
  - network
authors:
  - name: S. Thomas Kelly
    email: "tom.kelly@postgrad.otago.ac.nz, tom.kelly@riken.jp"
    orcid: 0000-0003-3904-6690
    affiliation: "1, 2" # (Multiple affiliations must be quoted)
  - name: Michael A. Black
    email: mik.black@otago.ac.nz
    orcid: 0000-0003-1174-6054
    affiliation: "1"
affiliations:
 - name: "Department of Biochemistry, University of Otago, PO Box 56, Dunedin 9054, New Zealand"
   index: 1
 - name: "RIKEN Center for Integrative Medical Sciences, Suehiro-cho-1-7-22, Tsurumi Ward, Yokohama, Kanagawa 230-0045, Japan"
   index: 2
date: "`r  format(Sys.time(), '%d %B %Y')`"
bibliography: paper.bib
header-includes:
  - \usepackage{caption}
---

@@

### Summary
Transcriptomic analysis is used to capture the molecular state of a cell
or sample in many biological and medical applications. In addition to 
identifying alterations in activity at the level of individual genes, 
understanding changes in the gene networks that regulate fundamental
biological mechanisms is also an important objective of molecular 
analysis. As a result, databases that describe biological pathways 
are increasingly uesad to assist with the interpretation of results
from large-scale genomics studies. Incorporating information from 
biological pathways and gene regulatory networks into a genomic data
analysis is a popular strategy, and there are many methods that provide
this functionality for gene expression data. When developing or comparing
such methods, it is important to gain an accurate assessment of their 
performance. Simulation-based validation studies are frequently used
for this.
This necessitates the use of simulated data that correctly accounts for
pathway relationships and correlations. Here we present a versatile
statistical framework to simulate correlated gene expression data from
biological pathways, by sampling from a multivariate normal distribution
derived from a graph structure. This procedure has been released as the
\texttt{graphsim} R package on CRAN and GitHub (\url{https://github.com/TomKellyGenetics/graphsim})
 and is compatible with any graph structure that can be described using
the `igraph` package. This package allows the simulation of biological
pathways from a graph structure based on a statistical model of gene expression.

@@

Network analysis of molecular biological pathways has the potential to
lead to new insights into biology and medical genetics
[@Barabasi2004; @Hu2016]. Since gene expression profiles capture a
consistent signature of the regulatory state of a cell
[@Perou2000; @Ozsolak2011; @Svensson2018], they can be used to analyse
complex molecular states with genome-scale data. However, biological
pathways are often analysed in a reductionist paradigm as amorphous sets
of genes involved in particular functions, despite the fact that the
relationships defined by pathway structure could further inform gene
expression analyses. In many cases, the pathway relationships are
well-defined, experimentally-validated, and are available in public
databases [@Reactome]. As a result, network analysis techniques can
play an important role in furthering our understanding of biological
pathways and aiding in the interpretation of genomics studies.

Gene networks provide insights into how cells are regulated, by mapping
regulatory interactions between target genes and transcription factors,
enhancers, and sites of epigenetic marks or chromatin structures
[@Barabasi2004; @Yamaguchi2007]. Inference using these regulatory
interactions genomic analysis has the potential to radically
expand the range of candidate biological pathways to be further
explored, or to improve the accuracy of bioinformatics and functional
genomic analysis. A number of methods have been developed to
utilise timecourse gene expression data [@Yamaguchi2007; @Arner2015]
using gene regulatory modules in state-space models and recursive vector
autoregressive models [@Hirose2008; @Shimamura2009]. Various approaches
to gene regulation and networks at the genome-wide scale have led to
novel biological insights [@Arner2015; @Komatsu2013], however, inference
of regulatory networks has thus far primarily relied on experimental
validation or resampling-based approaches to estimate the likelihood
of specific network modules being predicted [@Markowetz2007; @Hawe2019].

Simulating datasets that account for pathway structure are of particular
interest for benchmarking regulatory network inference techniques
and methods being developed for genomics data containing complex biological 
interactions [@Schaffter2011; @Saelens2019]. Dynamical models using
differential equations have been employed, such as by GeneNetWeaver
[@Schaffter2011], to generate simulated datasets
specifically for benchmarking gene regulatory network inference techniques.
There is also renewed interest in modelling biological pathways and simulating
data for benchmarking due to the emergence of single-cell genomics
technologies and the growing number of bioinformatics techniques developed
to use this data [@Zappia2017; @Saelens2019]. Packages such as 'splatter'
[@Zappia2017], which uses the gamma-poisson distribution,
have been developed to model single-cell data.
SERGIO [@Dibaeinia2019] and dyngen [@Cannoodt2020] build on
this by adding gene regulatory networks and multimodality
respectively. These methods have been designed based on known
deterministic relationships or synthetic reaction states,
to which stochasticity is then added.
However, it is computationally-intensive to model
these reactions at scale or run many iterations for benchmarking.
In some cases, it is only necessary to model the statistical
variability and "noise" of RNA-Seq data in order to evaluate
methods in the presence of multivariate correlation structures.

There is a need, therefore, for a systematic framework for statistical
modelling and simulation of gene expression data derived from
hypothetical, inferred or known gene networks. Here we present a
package to achieve this, where samples from a multivariate normal
distribution are used to generate normally-distributed log-expression
data, with correlations between genes derived from the structure of the
underlying pathway or gene regulatory network. This methodology enables
simulation of expression profiles that approximate the log-transformed
and normalised data from microarray studies, as well as, bulk or single-cell RNA-Seq
experiments. This procedure has been released as the \texttt{graphsim}
package to enable the generation of simulated gene expression datasets
containing pathway relationships from a known underlying network.
These simulated datasets can be used to evaluate various bioinformatics
methodologies, including
statistical and network inference procedures.

@@

Here we present a procedure to simulate gene expression data with
correlation structure derived from a known graph structure. This
procedure assumes that transcriptomic data have been generated and
follow a log-normal distribution (i.e.,
$log(X_{ij}) \sim MVN({\bf\mu}, \Sigma)$, where ${\bf\mu}$ and $\Sigma$
are the mean vector and variance-covariance matrix respectively, for
gene expression data derived from a biological pathway) after
appropriate normalisation [@Law2014; @Li2015]. Log-normality of gene
expression matches the assumptions of the popular \texttt{limma} package [@limma], which is
often used for the analysis of intensity-based data from gene expression
microarray studies and count-based data from RNA-Seq experiments. This
approach has also been applied for modelling UMI-based count data from
single-cell RNA-Seq experiments in the \texttt{DESCEND} R package [@Wang2018].

@@

Computing the nearest positive definite matrix is necessary to ensure
that the variance-covariance matrix can be inverted when used as a
parameter in multivariate normal simulations, particularly when negative
correlations are included for inhibitions (as shown below). Matrices
that cannot be inverted occur rarely with biologically plausible
graph structures but this approach allows for the computation of a
plausible correlation matrix when the given graph structure is
incomplete or contains loops. When required, the nearest positive
definite matrix is computed using the `nearPD` function of
the \texttt{Matrix} R package [@Matrix] to perform
Higham's algorithm [@Higham2002] on variance-covariance matrices.
The \texttt{graphsim} package gives a warning when this occurs.

@@

The graph structure in
Figure [1a](#fig:simple_graph:first){reference-type="ref"
reference="fig:simple_graph:first"} was used to simulate correlated gene
expression data by sampling from a multivariate normal distribution
using the \texttt{mvtnorm} R package [@Genz2009; @mvtnorm]. The graph structure
visualisation in
Figure [1](#fig:simple_graph){reference-type="ref"
reference="fig:simple_graph"} was specifically developed for (directed)
\texttt{igraph} objects in and is available in the \texttt{graphsim} package. The
\texttt{plot\_directed} function enables customisation of plot parameters for
each node or edge, and mixed (directed) edge types for indicating
activation or inhibition. These inhibition links (which occur frequently
in biological pathways) are demonstrated in
Figure [1b](#fig:simple_graph:second){reference-type="ref"
reference="fig:simple_graph:second"}.

@@

```{r, eval = FALSE}
#install packages required (once per computer)
install.packages("graphsim")
```
```{r, warning=FALSE, results='hide', message=FALSE}
#load required packages (once per R instance)
library("graphsim")
#load packages for examples
library("igraph"); library("gplots"); library("scales")
```

@@

```{r simple_graph, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.375\\linewidth', out.width='.375\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulated graph structures}. A constructed graph structure used as an example to demonstrate the simulation procedure in Figures 2 and 3. Activating links are denoted by black arrows and inhibiting links by red edges.', echo=FALSE}
#generate graph structure
graph_edges <- rbind(c("A", "C"), c("B", "C"), c("C", "D"), c("D", "E"),
   c("D", "F"), c("F", "G"), c("F", "I"), c("H", "I"))
graph <- graph.edgelist(graph_edges, directed = TRUE)

@@

#simulate data
expr <- generate_expression(100, graph, cor = 0.8, mean = 0,
  comm = FALSE, dist =TRUE, absolute = FALSE, state = state)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none",
          col = colorpanel(50, "white", "red"),
  colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
  colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
```

```{r simulation_activating, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.375\\linewidth', out.width='.375\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from a graph structure}. An example of a graph structure (a) that has been used to derive a relationship matrix (b), $\\Sigma$  matrix (c) and correlation structure (d) from the relative distances between the nodes. Non-negative values are coloured white to red from $0$ to $1$ (e). The $\\Sigma$ matrix has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here genes with closer relationships in the pathway structure show a higher correlation between simulated values.', warning=FALSE, echo=FALSE}
# activating graph
state <- rep(1, length(E(graph)))
plot_directed(graph, state=state, layout = layout.kamada.kawai,
  cex.node=2, cex.arrow=4, arrow_clip = 0.2)
mtext(text = "(a) Activating pathway structure", side=1, line=3.5, at=0.075, adj=0.5, cex=1.75)
box()
#plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
  scale = "none", trace = "none", key = FALSE,
  col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(b) Relationship matrix", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, cor = 0.8, absolute = FALSE),
scale = "none", trace = "none", key = FALSE,
col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#simulate data
expr <- generate_expression(100, graph, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE, state = state)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", key = FALSE, 
          col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(d) Simulated correlation", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", key = FALSE, col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "samples", side=1, line=1.5, at=0.2, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=-0.4, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
```

```{r, include=FALSE, warning=FALSE}
pdf("Plotsimulation_activating-5.pdf", width = 12.14, height = 6.072)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", key = FALSE, col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "", mar = c(6, 6), keysize = 1)
mtext(text = "samples", side=1, line=1.5, at=0.55, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=0.45, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0.5, adj=0.5, cex=1.75)
dev.off()
#system("sed -i.bak '/Plotsimulation_activating-5/s/\\(.*\\)width=[.]415/\1width=.830/g' paper.md && rm paper.md.bak")
#system("sed -i.bak '/Plotsimulation_activating-5/s/\\(.*\\)width=[.]415/\1width=.830/g' paper.tex && rm paper.tex.bak")
#sed -i '/Plot.*-5/s/\(.*\)width=[.]415/\1width=.830/g' paper.md paper.tex
```

@@

```{r simulation_inhibiting, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.375\\linewidth', out.width='.375\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from graph structure with inhibitions}. An example of a graph structure (a), that has been used to derive a relationship matrix (b), $\\Sigma$ matrix (c), and correlation structure (d), from the relative distances between the nodes (e). These values are coloured blue to red from $-1$ to $1$. This has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here the inhibitory relationships between genes are reflected in negatively correlated simulated  values.', warning=FALSE, echo=FALSE}
#generate parameters for inhibitions
state <- c(1, 1, -1, 1, 1, 1, 1, -1)
plot_directed(graph, state=state, layout = layout.kamada.kawai,
  cex.node=2, cex.arrow=4, arrow_clip = 0.2)
mtext(text = "(a) Inhibiting pathway structure", side=1, line=3.5, at=0.075, adj=0.5, cex=1.75)
box()
#plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
  scale = "none", trace = "none", key = FALSE, col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(b) Relationship matrix", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, state, cor = 0.8, absolute = FALSE),
scale = "none", trace = "none", key = FALSE, col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#simulated data
expr <- generate_expression(100, graph, state, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE)
#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", key = FALSE, col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(d) Simulated correlation", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", key = FALSE, col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "")
mtext(text = "samples", side=1, line=1.5, at=0.2, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=-0.4, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0, adj=0.5, cex=1.75)
```

```{r, include=FALSE, warning=FALSE}
pdf("Plotsimulation_inhibiting-5.pdf", width = 12.14, height = 6.072)
#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", key = FALSE, col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)), labCol = "", mar = c(6, 6), keysize = 1)
mtext(text = "samples", side=1, line=1.5, at=0.55, adj=0.5, cex=1.5)
mtext(text = "genes", side=4, line=1, at=0.45, adj=0.5, cex=1.5)
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3.5, at=0.5, adj=0.5, cex=1.75)
dev.off()
#("sed -i.bak '/Plotsimulation_inhibiting-5/s/\\(.*\\)width=[.]415/\1width=.830/g' paper.md && rm paper.md.bak")
#system("sed -i.bak '/Plotsimulation_inhibiting-5/s/\\(.*\\)width=[.]415/\1width=.830/g' paper.tex && rm paper.tex.bak")
#sed -i '/Plot.*-5/s/\(.*\)width=[.]415/\1width=.830/g' paper.md paper.tex
```
